---
title: "Introduction to regkit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to regkit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE}
library(regkit)
log_file <- tempfile()
cat("Vignette log file", file = log_file)
```

**regkit** aims to facilitate the manipulation, analysis and visualization of data from Norwegian health and population registers and capitalize on their characteristics.

Researchers with access to microdata from health and/or administrative registers in Norway can use **regkit** to streamline their initial analytical processes. The package includes functions for data pre-processing, linkage, and the computation of relevant statistics and visualizations, such as stratified frequencies, incidence and prevalence rates.

This vignette introduces **regkit**'s main functions and shows examples of how to use them with individual-level data. The package also includes some helper functions that can aid researchers working with registry data in Norway address specific problems, more information in the `vignette("h-functions")` and `vignette("other-useful-fun")`.

## Example datasets

To exemplify the main functions of the package, we have included a couple of illustrative simulated datasets. These datasets are also used in the examples specified in the functions' documentation.

The dataset `diag_df` is a tibble with simulated individual-level diagnostic data. It is documented in `?diag_df`

```{r diagdf, echo=TRUE}
str(diag_df)
```

The dataset `var_df` is a tibble with simulated individual-level time-varying data. It is documented in `?var_df`

```{r vardf, echo=TRUE}
str(var_df)
```

The dataset `invar_df` is a tibble with simulated individual-level time-invariant data. It is documented in `?invar_df`

```{r invardf, echo=TRUE}
str(invar_df)
```

All the datasets above have been created using the `synthetic_dataset()` function included in this package. The exact code used to create them can be found in the folder `/data-raw` in the package's source code. The `vignette("h_functions")` includes more detailed information about simulating datasets in `regkit`.

## Data reading and validation

Generally, researchers using Norwegian individual-level registry data for epidemiological research will access datasets that fall into one of the following three categories:

-   **Diagnostic data:** at least including unique personal identifier (ID), date of diagnostic event, and diagnosis code (such as ICD-10 or ICPC-2). For instance, datasets from NPR (Norwegian Patient Registry) or KPR (Kommunalt pasient- og brukerregister).

-   **Time-invariant administrative data:** administrative sociodemographic data that does not change for a given individual in the population such as date of birth, immigration background, etc.

-   **Time-varying administrative data:** administrative sociodemographic data that can change for a given individual in the population, such as place of residence or marital status. This type of information usually is updated quarterly or yearly in administrative registries.

Assuming a researcher would want to compute prevalence rates for a certain diagnosis stratified by some sociodemographic characteristic(s), they would at least need a *diagnostic* dataset and at least one administrative dataset.

The functions `read_diag_data()` and `read_admin_data()` check the datasets' minimum requirements given the users expectations about what they contain, read them into memory, and give a quick summary of their structure. For example, in the case of a dataset the user has identified as containing *time-invariant* data, it is assumed that each row in the dataset corresponds to a individual identified by a unique personal identifier. To allow the package to check that the dataset meets minimum requirements for being useable *time-invariant* data, it is necessary for the user to specify some of the column names - i.e., corresponding to the id or date variables (see the function documentation for more info).

<br>

```{=html}
<style>
div.blue { border: 4px #65699d; border-style: none none none solid; padding: 3px;}
</style>
```

::: blue
📝 **Note**: these functions only check for *minimum* requirements; datasets may contain additional variables/columns. For example, *diagnostic data* sometimes includes sex and age information for each individual, even though these are not minimally required for this data type.
:::

<br>

The `read_..._data()` functions support common file formats that researchers might realistically encounter in data deliveries, such as CSV, SAV, and *parquet* files. Parquet files are specially useful when working with very large data, as the storage and data processing is more efficient than other types of file formats (e.g. CSV files). As explained in the `vignette("parquet-files")`, using parquet files is recommended for large and larger-than-memory data within **regkit**. The vignette also includes general instructions on how to write/save parquet datasets.

### Read example diagnostic data

To read and validate the minimum requirements of a CSV file containing *diagnostic data*:

```{r read-diag, echo=TRUE}
diag_csv <- system.file("extdata", "diag_data.csv", package = "regkit")

diag_data_validated <- regkit::read_diag_data(
  diag_csv,
  id_col = "id",
  date_col = "diag_year",
  log_path = log_file
)
```

### Read example administrative data

Similarly, to read and validate the minimum requirements of a CSV file containing *time invariant data*:

```{r read-demo, echo=TRUE}
demo_csv <- system.file("extdata", "invar_data.csv", package = "regkit")

demo_data_validated <- read_admin_data(
  demo_csv,
  data_type = "t_invariant",
  id_col = "id",
  log_path = log_file
)
```

In the case of CSV, RDS/RDA, and SAV files, the resulting output is a tibble that can be further passed as input to other functions in the package. For parquet files/datasets, the output is an ArrowObject that can be used as input in the filtering functions described in the section below. For more information about ArrowObject, please consult the documentation from the package `arrow`.

## Data filtering

Usually, individual-level register data has a large number of observations which makes it cumbersome to manipulate and prepare for analysis. For that reason, it is advantageous to follow a 'filter first' approach and remove non-relevant variables and observations.

As mentioned in the `vignette("parquet-files")`, parquet files are more efficient than other type of files at performing operations on large datasets. Therefore, both the `filter_diag_data()` and `filter_admin_data()` functions are more memory efficient when providing an ArrowObject as an input (see section above).

Regardless of the input's data type (ArrowObject or tibble), the output of these functions are tibbles. It is assumed that after the initial filtering of both diagnostic and sociodemographic data, the datasets will be smaller and easier for users to manipulate as in-memory tibbles.

::: lightblue
ℹ️ **Variables and codes?**

In order to correctly use the `filter_..._data` functions, it is necessary to have previous knowledge of the data sources and variables used. In most cases, data deliveries from registry sources are accompanied with metadata documentation for each dataset. In the case that the specific project metadata is not available, here are some useful starting points:

-   **SSB**: [Statistical Classifications and Codelists (Klass)](https://www.ssb.no/klass/), [Metadata](https://www.ssb.no/a/metadata/), [Variable lists](https://www.ssb.no/data-til-forskning/utlan-av-data-til-forskere/variabellister).

-   **Health data**: [Helsedata](https://helsedata.no/no/), [FinnKode](https://finnkode.helsedirektoratet.no), [Norwegian Institute of Public Health](https://www.fhi.no/he/)
:::

### Filter diagnostic data

Due to the distinct characteristics of *diagnostic* and *time-varying/time-invariant* datasets, there are two filtering functions: `filter_diag_data()` and `filter_admin_data()`. The former checks that the ICD-10 or ICPC-2 codes/family or codes given by the user (`codes` argument) are currently codes reported to [NPR](https://finnkode.helsedirektoratet.no/icd10/chapter) or [KPR](https://finnkode.helsedirektoratet.no/icpc2/chapter). Afterwards, it filters the given diagnostic dataset to keep only the observations with the relevant ICD-10 or ICPC-2 codes. Additionally, it is possible to filter by date of the diagnostic event and remove all rows with missing data.

```{=html}
<style>
div.red { border: 4px #f95d6d; border-style: none none none solid; padding: 3px;}
</style>
```

::: red
⚠️ Sometimes the diagnostic data might contain some subcodes not of mandated reporting to NPR or KPR, therefore the `filter_diag_data()` throws a warning instead of an error when the codes given by the user are not found in the list of current valid and reported codes to NPR or KPR. In that case, it is **important** to verify the validity of the exact codes given in alternative sources, such as [FinnKode](https://finnkode.helsedirektoratet.no).
:::

For example, to keep only the observations that either have the ICD-10 code *F840* or *F841*, between the years of 2016 and 2017:

```{r filter-diag-codes, echo = TRUE}
filtered_diag_codes <- filter_diag_data(
  data = diag_df,
  codes = c("F840", "F841"),
  classification = "icd",
  id_col = "id",
  code_col = "code",
  date_col = "diag_year",
  diag_dates = c("2016", "2017"),
  log_path = log_file
)
```

Alternatively, it is also possible to filter by pattern/family of codes. For example, to keep the observations with all valid ICD-10 codes starting with *F45* or *F84* use the argument `pattern_codes` instead of `codes`. Additionally, the `add_description` argument adds a column containing a short description of the diagnostic code.

```{r filter-diag-pattern, echo = TRUE}
filtered_diag_pattern <- filter_diag_data(
  data = diag_df,
  pattern_codes = c("F45", "F84"),
  classification = "icd",
  id_col = "id",
  code_col = "code",
  add_description = TRUE,
  log_path = log_file
)
```

<br>

```{=html}
<style>
div.lightblue {background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
```

::: lightblue
ℹ️ **Curate diagnostic data**

For some analyses, it will be necessary to further filter the diagnostic dataset. For example, if one is interested in the age of first diagnosis for a chronic disease, only the first registered diagnosis information is relevant. Or, sometimes it is desirable to only keep cases with more than a certain number of registered diagnostic events. To accomplish this type of summarization and curation of the diagnostic dataset, the `curate_diag_data()` provides some additional filtering options. It is important to highlight that this function only supports data frames (preferably tibbles) as input.

For example, to summarize information by first-time diagnosis:

```{r curate-diag, echo = TRUE}
curated_diag <- curate_diag_data(
  data = filtered_diag_pattern,
  min_diag = 1,
  first_diag = TRUE,
  id_col = "id",
  code_col = "code",
  date_col = "diag_year",
  log_path = log_file
)
```
:::

### Filter time-varying and time-invariant data

Similar to the filtering of diagnostic data, `filter_admin_data()` aids with the filtering of both time-varying and time-invariant datasets.

For example, to only keep observations where the individuals have resided in the municipality "0815" between the years 2012 and 2015:

```{r filter-var, echo = TRUE}
filtered_var <- filter_admin_data(
  data = var_df,
  data_type = "t_variant",
  filter_param = list("year_varying" = c(2012:2015), "varying_code" = c("0815")),
  log_path = log_file
)
```

In the case of time-invariant data, to keep only individuals with year of birth between 2010-2018 and reason of immigration "ARB", "NRD" or "UKJ":

```{r filter-invar, echo = TRUE}
filtered_invar <- filter_admin_data(
  data = invar_df, data_type = "t_invariant",
  filter_param = list("y_birth" = c(2010:2018), "innvandringsgrunn" = c("ARB", "NRD", "UKJ")),
  rm_na = FALSE,
  log_path = log_file
)
```

## Linkage

In order to use information from datasets containing *diagnostic* and *time-varying* or *time invariant* data in analyses (e.g., the calculation of stratified prevalence rates), it is necessary to link them. In the **regkit** workflow, linkage using the individuals' unique personal identifiers present in each dataset comes after filtering, to save on memory and reduce the potential for merging errors.

<br>

::: blue
📝 **Note**: depending on the type of analysis and relevant variables, it *might* not be necessary to link any datasets, in which case this step can be skipped.
:::

<br>

The `link_diag_admin()` function can aid with the linkage process, as long as all datasets share the same IDs to uniquely identify individuals. To link both the already filtered diagnostic and time-invariant datasets:

```{r link-data, echo = TRUE}
linked_diag_inv <- link_diag_admin(
  data_diag = curated_diag,
  data_admin_inv = filtered_invar,
  id_col = "id",
  log_path = log_file
)
```

As part of the linking process, the function provides some useful information about the number of matched cases (individuals) and summary information of the new linked dataset. The resulting dataset from `link_diag_admin()` is a minimal dataset that contains all relevant observations to perform further analyses.

## Analysis

Until this point, all the functions described are concerned with general data preparation and manipulation that researchers working with this type of individual-level data have to perform. The functions described in the next section are specific to descriptive epidemiology and public health monitoring - though elements of them are generalizable to other contexts.

Two of the most common statistics in descriptive epidemiology are incidence and prevalence rates. Consequently, **regkit** includes the functions `calculate_prevalence()` and `calculate_incidence()` to aid with these analyses. Both functions require as input a filtered linked dataset with relevant observations (output from `link_diag_admin()`), and population counts stored in a different dataset.

Both functions internally operate in the same way, it is then crucial to verify that the linked dataset contains only relevant observations for the desired analysis. Briefly, incidence refers to the number of **new cases** of a disease that occur in a specified population during a period of time. While, prevalence describes the total number of **existing cases** in the population at a specific point or period in time. In order to then correctly calculate incidence rates or incidence proportions, the linked dataset should only contain **new cases** in our population of interest. In contrast, to calculate prevalence rates the linked dataset should include all existing (both old and new) cases in our population.

<br>

::: lightblue
ℹ️ **Population**

In order to correctly compute the incidence and prevalence, it is necessary to know or calculate the relevant population counts, which will differ depending on the exact measure one is interested. Below you can find a brief summary of some common definitions used in epidemiology:

| Concept | Definition | Population (denominator) |
|------------------------|------------------------|------------------------|
| Incidence rate | Rate at which new cases occur over time | Total person-time at risk: sum of each time each person in the population is at risk of developing the disease |
| Incidence proportion or cumulative incidence | Proportion of new cases over a period of time | Disease-free individuals at the start of the period of interest |
| Prevalence rate | Proportion of existing cases at a point or period of time | Everyone in the population (with or without the disease) |

As the definition and calculation of population will greatly vary depending on each research question and desired analyses, **regkit** currently does not include any specific functions to compute population counts from individual-level data. However, in most cases the study population should be available as metadata to the research project. Otherwise, in the case of having individual-level data with **full coverage** of the population of interest, it is possible to compute population counts based on that.

For publicly available population counts from SSB, the function `get_population_ssb()` can retrieve population counts by year, age, sex and place of residence (municipality and county level).
:::

<br>

As diagnoses under F84 and F45 are considered to be chronic or persistent, it is assumed for the next examples that once an individual is diagnosed, then they will always be part of the case group. Then, to compute the prevalence of F84 and F45 (and subcodes) given in the time period 2012-2020 to individuals born between 2010-2018 and with reason of immigration "ARB", "NRD" or "UKJ":

```{r prevalence, echo = TRUE}

pop_df <- tibble::tibble(year = "2012-2020", population = 30024)
linked_diag_inv <- linked_diag_inv |> dplyr::rename("year"= "y_diagnosis_first")

prevalence_df <- calculate_prevalence(linked_diag_inv,
  id_col = "id",
  date_col = "year",
  pop_data = pop_df,
  pop_col = "population",
  time_p = c(2012,2020),
  CI = TRUE,
  CI_level = 0.95,
  only_counts = FALSE,
  suppression = TRUE,
  suppression_threshold = 10,
  log_path = log_file)

prevalence_df

```

Besides computing prevalence rates based on the provided linked dataset, `calculate_prevalence()` can also provide confidence intervals, and can suppress of low counts (this is done by default) to help researchers with responsible reporting of results. Furthermore, there is an option to output only the case counts, which can be useful for other type of statistical analyses (e.g. Chi square tests).

Often, it is relevant to compute counts or rates stratified by certain groupings. It is also possible to specify this, as long as the population dataset includes the necessary information (with a warning if it does not have a one-to-one match):

```{r prevalence-strat, echo = TRUE}
pop_df <- tidyr::expand_grid(year = "2012-2020",
                               sex = as.factor(c(0, 1)),
                               innvandringsgrunn = c("ARB", "UKJ", "NRD")) |>
    dplyr::mutate(population = floor(runif(dplyr::n(), min = 3000, max = 4000)))


prevalence_df_strat <- calculate_prevalence(linked_diag_inv,
  id_col = "id",
  date_col = "year",
  pop_data = pop_df,
  pop_col = "population",
  grouping_vars = c("sex","innvandringsgrunn"),
  time_p = c(2012,2020),
  CI = TRUE,
  CI_level = 0.95,
  only_counts = FALSE,
  suppression = TRUE,
  suppression_threshold = 10,
  log_path = log_file)

prevalence_df_strat

```

Often in descriptive epidemiology, researchers are interested in the evolution of prevalence of a certain disease through time. For that purpose, the `calculate_prevalence_series()` can compute prevalence rates for a time series. The `calculate_incidence_series()` follows the same logic.

```{r prevalence-series, echo = T, results = 'hide', message=FALSE}

# Silenced CLI output for example 

pop_df <- tidyr::expand_grid(year = c("2012-2014", "2015-2017", "2018-2020"),
                             sex = as.factor(c(0, 1)),
                             innvandringsgrunn = c("ARB", "UKJ", "NRD")) |>
  dplyr::mutate(population = floor(runif(dplyr::n(), min = 2000, max = 4000)))

prevalence_df_series <- calculate_prevalence_series(linked_diag_inv,
  time_points = list(c(2012,2014), c(2015,2017), c(2018,2020)),
  id_col = "id",
  date_col = "year",
  pop_data = pop_df,
  pop_col = "population",
  grouping_vars = c("sex", "innvandringsgrunn"),
  only_counts = FALSE,
  suppression = TRUE,
  suppression_threshold = 1,
  CI = TRUE,
  CI_level = 0.95,
  log_path = log_file)

```

```{r, echo= FALSE}
prevalence_df_series
```

## Visualize

In the case of comparison between different groups or changes through time, it is useful to visualize the prevalence/incidence rates. For that purpose, the function `plot_rates()` can create some ready-to-use plots with a consistent visual theme:

```{r plot-rates, fig.width=8, fig.height=6, dpi = 120,  out.width="80%"}

plot_line <- plot_rates(prevalence_df_series,
                        date_col = "year",
                        rate_col = "prev_rate",
                        plot_type = "line",
                        grouping_var = "innvandringsgrunn",
                        facet_var = "sex",
                        palette = "fhi_colors",
                        CI_lower = "ci_results_lower",
                        CI_upper = "ci_results_upper",
                        plot_title = "Prevalence by sex and reason of immigration",
                        x_name = "Year",
                        start_end_points = TRUE)

plot_line

```

As the output of this function is a ggplot object, it is possible to further modify using the `ggplot2` suite of functions.

```{r, fig.width=8, fig.height=6, dpi = 120, out.width="80%"}

plot_line + ggplot2::labs(subtitle = "Norway, individuals born between 2010-2018")
  
```

## Logging

To facilitate reproducibility and transparency across research projects, the majority of functions in **regkit** create log files that document their internal data processing, warnings/errors and corresponding outputs. Similarly, each function provides clear console feedback after it executes important operations (filter, select, join, etc).

It is possible to either provide the path to an already existing .log file or set the argument `log_path = NULL` (default) to create a new /log directory and .log file for each function. For instance, the `read_diag_data()` with a `NULL` log_path will first check in the working directory for a `/log` directory. In case it does not already exist, the `/log` directory is created in the current working directory and a \<read_diag_data_dd_mm_yyyy.log\> file is initialized.

## Other useful functions

In addition to the main functions in the **regkit** workflow, the package includes some additional functions that can be helpful in the process of working with Norwegian individual-level registry data. Please refer to the specific vignettes for more information: `vignette("other-useful-fun")`
