---
title: "Reference for package dependencies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reference for package dependencies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(regkit)
```

The package `regkit` has the following Imports and Suggests:


```{r dep, echo = F}

desc::desc_get_deps(".")

```

Each of those packages has a number of dependencies (recursive dependencies): 


```{r recursive, echo = F}

direct <- desc::desc_get_deps(".")$package
direct <- setdiff(direct, "R")

recursive_deps <- tools::package_dependencies(
  direct,
  which = c("Depends", "Imports"),    
  recursive = TRUE
)

all_deps <- sort(unique(unlist(recursive_deps)))
all_deps

```


### Dependency tree 


You can also visualize how all the dependencies fit together: 

```{r dep-tree, echo = F}

 pak::pkg_deps_tree(".")

```


## Installing dependencies in secure environments 

The package `regkit` is meant to be used inside a secure computing environment with limited internet access, thus some dependencies might have to be installed manually. One solution is to create a miniCRAN repository containing all dependencies necessary to successfully install `regkit`. 

As a reference, here is the workflow to build a miniCRAN repository assuming the secure environment you are installing it runs Windows: 

1. Install the `miniCRAN` package and `regkit` outside the secure environment 
2. Save the dependencies using the following code: 



```{r get-deps, eval = F}

library(miniCRAN)
pkg_list <- pkgDep("regkit", repos = "https://cloud.r-project.org", type = "source", suggests = TRUE)

makeRepo(pkg_list, path = "/path/to/myrepo", repos = "https://cloud.r-project.org", type = "source")
saveRDS(pkg_list, "pkg_list.rds")

```

3. Import the miniCRAN folder and dependency list `pkg_list.rds` to the secure environment 
4. Read `pkg_list.rds` and install dependencies from miniCRAN inside secure environment: 

```{r install-deps, eval = F}

pkg_list <- readRDS("pkg_list.rds")

install.packages(
  pkg_list,  
  repos = paste0("file:///", "/path/to/myrepo"),
  type = "source"
)

```

4. Install `regkit` from the tarball you can download from the [Releases
section](https://github.com/amslala/regkit/releases) in GitHub: 


```{r install-pkg, eval = F}

pkg_list <- readRDS("pkg_list.rds")

install.packages(
  "/path/to/regkit_x.y.z.tar.gz",  
  repos = NULL,
  type = "source"
)

```






